---
title: "Employee Attrition Analaysis"
author: "Pearly Merin Paul"
date: "8/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction 

In the below report, we have been provided with the  data set `CaseStudy.csv` to analyze the factors that lead to attrition. Our team in DDS Analytics were asked to identify the top three factors that contribute to turnover .
Our team had to do an exploratory Data Analysis on the data provided to come up with useful insights.
Along with the base data we were provided datasheet `CaseStudy2CompSet No Attrition` 
to use the models to predict Attrition

And an another datasheet `CaseStudy2CompSet No Salary.csv` to predict the Monthly Income of the Employee.

## Reproducible Research in R
​
Please begin by reading the the README.md.All the details on the structure of project has been documented.
​
### Loading Packages for Analysis
Below are the packages required for the analysis of the Employee Attrition Analyis 

```{r Packages}
library(magrittr)
library(dplyr)
library(ggplot2)
library(Amelia) # Confirm missing value
library(rpart)
library(rattle)
require(caTools)
library(ggcorrplot)
library(randomForest)
library(randomForestSRC)
library(corrplot)
library(cowplot)
library(class)
library(caret)

```
## Loading all the datasheets for analysis 

`CaseStudy.csv` : To perform exploratory Data analysis 
`CaseStudy2CompSetNoSalary.csv` : To predict Salary
`CaseStudy2CompSet No Attrition.csv`: To predict Attrition


```{r}

employee <- read.csv("CaseStudy.csv")

employeeNoAttrition<-read.csv("data/CaseStudy2CompSet No Attrition.csv")
employeeNoSalary<-read.csv("data/CaseStudy2CompSetNoSalary.csv")

employeedata<-employee
 employee <- select(employee,-c(ID, EmployeeCount, EmployeeNumber, Over18,StandardHours))

employeetotal<-employee

glimpse(employeetotal)
employeetotal1<-filter(employeetotal,Attrition=="Yes")


 
```

## Overall Attrition Analysis

### We are able to see the data seemed to be not even . MEaning the base data sheet that we were able to that percentage of Employees who were attrition was 83% and the percentage of the employees who were not was 16.1%
```{r}

dt <- employeedata%>%
  dplyr::group_by(Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/sum(n))

theme_set(theme_bw())  # pre-set the bw theme.
 ggplot(data = dt,aes(x= Attrition, y = n))+ geom_bar(stat="identity",fill=c("blue","purple"))+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=.8), colour="white", size = 5)+theme(axis.text.x = element_text(angle = 90,hjust =0 )) + 
   ggtitle("Overall Attrition")+ylab("Count Of Employees")
```

## Cleaning Of the data

In this chunk we have done the data cleaning . HEre were have take the quantitative variables in the data and converted them into categorical variable by this we can get a better model and do exploratory data anlysis on the data.
```{r}


employeedata<-employee

employeedata$Age <- cut(employeedata$Age,breaks=c(-Inf,20, 30, 40, 50, 60,Inf),labels=c("0-19 years","20-29 years","30-39 years","40-49 years","50-59 years","60 +"),right=FALSE)


employeedata$DailyRate <- cut(employeedata$DailyRate,breaks=c(-Inf,500, 1000,1500, 2000,Inf),labels=c("0-499 Range","500-999 Range","1000-1499 Range","1500 + Range","60 +"),right=FALSE)


employeedata$DistanceFromHome <- cut(employeedata$DistanceFromHome,breaks=c(-Inf,10, 20,30,Inf),labels=c("1-9 miles","20-29 miles","30-40 miles","40+ miles" ),right=FALSE)


employeedata$MonthlyIncome <- cut(employeedata$MonthlyIncome,breaks=c(-Inf,5000, 10000,20000,Inf),labels=c("Low","Medium","Higher","Highest" ),right=FALSE)



employeedata$HourlyRate <- cut(employeedata$HourlyRate,breaks=c(-Inf,50,100, Inf),labels=c("0-150 dollars","50-100 dollars","100+ dollars" ),right=FALSE)




employeedata$MonthlyRate <- cut(employeedata$MonthlyRate,breaks=c(-Inf,5000, 10000,20000,Inf),labels=c("Low","Medium","Higher","Highest" ),right=FALSE)

employeedata$YearsAtCompany <- cut(employeedata$YearsAtCompany,breaks=c(-Inf, 10,15,Inf),labels=c("1-9 years","10 -14 years","15 + years" ),right=FALSE)

employeedata$TotalWorkingYears <- cut(employeedata$TotalWorkingYears,breaks=c(-Inf, 10,15,Inf),labels=c("1-9 years","10 -14 years","15 + years" ),right=FALSE)






employeedata$Education=as.factor(if_else(employeedata$Education == 1,"Below College", if_else(employeedata$Education == 2, "College", if_else(employeedata$Education == 3, "Bachelor", if_else(employeedata$Education == 4, "Master","Doctor")))))

employeedata$EnvironmentSatisfaction = as.factor(if_else(employeedata$EnvironmentSatisfaction == 1,"Low",if_else(employeedata$EnvironmentSatisfaction == 2, "Medium", if_else(employeedata$EnvironmentSatisfaction == 3, "High", "Very High"))))

employeedata$JobInvolvement = as.factor(if_else(employeedata$JobInvolvement == 1,"Low",if_else(employeedata$JobInvolvement == 2, "Medium",if_else(employeedata$JobInvolvement == 3, "High", "Very High"))))


employeedata$JobSatisfaction = as.factor(if_else(employeedata$JobSatisfaction == 1, "Low",if_else(employeedata$JobSatisfaction == 2, "Medium",if_else(employeedata$JobSatisfaction == 3, "High","Very High"))))

employeedata$PerformanceRating = as.factor(if_else(employeedata$PerformanceRating == 1, "Low",if_else(employeedata$PerformanceRating == 2, "Good", if_else(employeedata$PerformanceRating == 3, "Excellent", "Outstanding"))))

employeedata$RelationshipSatisfaction = as.factor(if_else(employeedata$RelationshipSatisfaction == 1, "Low",if_else(employeedata$RelationshipSatisfaction == 2, "Medium", if_else(employeedata$RelationshipSatisfaction == 3, "High", "Very High"))))

 employeedata$WorkLifeBalance = as.factor(if_else(employeedata$WorkLifeBalance == 1, "Bad",if_else(employeedata$WorkLifeBalance == 2, "Good", if_else(employeedata$WorkLifeBalance == 3, "Better", "Best"))))
 
 
 
employeedata$JobLevel = as.factor(if_else(employeedata$JobLevel == 1, "Low",if_else(employeedata$JobLevel == 2, "Medium",if_else(employeedata$JobLevel  == 3, "High","Very High"))))


employeedata$YearsSinceLastPromotion = cut(employeedata$YearsSinceLastPromotion ,breaks=c(-Inf, 10,15,Inf),labels=c("1-9 years","10 -14 years","15 + years" ),right=FALSE)

employeedata$PercentSalaryHike  = cut(employeedata$PercentSalaryHike  ,breaks=c(-Inf, 10,15,20,Inf),labels=c("1-9 percent","10 -14 percent","15 + percent","20+ percent" ),right=FALSE)

employeedata$YearsInCurrentRole= cut(employeedata$YearsInCurrentRole ,breaks=c(-Inf, 10,15,20,Inf),labels=c("1-9 years","10 -14 years","15 + years","20+ years" ),right=FALSE)


employeedata$YearsWithCurrManager= cut(employeedata$YearsWithCurrManager ,breaks=c(-Inf, 10,15,20,Inf),labels=c("1-9 years","10 -14 years","15 + years","20+ years" ),right=FALSE)



employeedata$NumCompaniesWorked= cut(employeedata$NumCompaniesWorked ,breaks=c(-Inf, 3,6,9,Inf),labels=c("0-2 years","3-5 years","7-8 years","9+ years" ),right=FALSE)

```

## Correlation Plot

Corelation plot between the various features provided in the datasheet.

Upin analysis of the Pvalues of the features we are able to see that 

JobLevel, OverTime and StockLevel Options
``{r}

employee_updated<-employee

employee_updated$OverTime=as.numeric(as.factor(employee_updated$OverTime))
employee_updated$Attrition=as.numeric(as.factor(employee_updated$Attrition))

employee_updated$MaritalStatus=as.numeric(as.factor(employee_updated$MaritalStatus))
employee_updated$BusinessTravel=as.numeric(as.factor(employee_updated$BusinessTravel))

employee_updated$Department=as.numeric(as.factor(employee_updated$Department))

employee_updated$JobRole=as.numeric(as.factor(employee_updated$JobRole))

employee_updated$EducationField=as.numeric(as.factor(employee_updated$EducationField))

employee_updated$Gender=as.numeric(as.factor(employee_updated$Gender))

employee_updated$StockOptionLevel=as.numeric(as.factor(employee_updated$StockOptionLevel))
corrplot( cor(as.matrix(employee_updated), method = "pearson", use = "complete.obs") ,is.corr = FALSE, type = "lower", order = "hclust", tl.col = "black", tl.srt = 360)
library(ggcorrplot)
corrplot(cor(employee_updated), order = "alphabet",method= "number", type = "full",
         col = brewer.pal(n = 8, name = "RdBu"))


ggcorrplot(cor(employee_updated),  hc.order = TRUE, type = "lower",method= "circle",
  outline.color = "black",
  ggtheme = ggplot2::theme_gray,
  colors = c("#6D9EC1", "white", "#E46726")
)

 library(psych)
 corr.test( employee_updated)
```

## Employee Attrition vs Environment Satisfaction
```{r}
dtEnv <- employeedata%>%
  dplyr::group_by(EnvironmentSatisfaction,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 ggplot(data = dtEnv ,aes(x=EnvironmentSatisfaction , y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))


```




## 1. What are the major factors that impact employee Attrition and  income benefits of an employee
Benefits of an employee play a major role in the employees attrition rate. An employee who is payed well tend to be more happy and stick with the company .

## Employee Attrition vs Stock Option Level
# Employees get either level 0 or 1 of the company stocks . Employees who were offered no stock opetions tend to quite the company. Giving stock option to an employee is beneficial. Because by giving stock options they are technically stake holders 



```{r}
dtStockLevel <- employeedata%>%
  dplyr::group_by(StockOptionLevel,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p1<-ggplot(data = dtStockLevel,aes(x=StockOptionLevel, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Stock Option -Attrition Analysis")+ylab("Count Of Employees")

 p1
```

## Employee Attrition vs Monthly Rate of Employee

```{r}
dtMonthlyRate <- employeedata%>%
  dplyr::group_by(MonthlyRate,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p2<-ggplot(data = dtMonthlyRate,aes(x=MonthlyRate, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Monthly -Attrition Analysis")+ylab("Count Of Employees")

p2
```

## Employee Attrition vs Hourly Rate of Employee

```{r}
dthour <- employeedata%>%
  dplyr::group_by(HourlyRate,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p3<-ggplot(data =dthour,aes(x=HourlyRate, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Hourly Rate -Attrition Analysis")+ylab("Count Of Employees")


p3
```


## Employee Attrition vs Monthly Income of an Employee

```{r}
dtIncome <- employeedata%>%
  dplyr::group_by(MonthlyIncome,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p5<- ggplot(data = dtIncome,aes(x= MonthlyIncome, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Monthly Income-Attrition Analysis")+ylab("Count Of Employees")
 
p5

```
## Employee Attrition vs Percentage of Salary Hike 

### Upon analysing the Percent Salary Hike versus employees with highest Attrition has 10-14 % percentage hike. Employees with less hike in percentage,

53.6% - falls in the 10-14% 


```{r}
dtHike <- employeedata%>%
  dplyr::group_by(PercentSalaryHike,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p6<- ggplot(data = dtHike ,aes(x= PercentSalaryHike, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Salary Hike -Attrition Analysis")+ylab("Count Of Employees")
 
p6

```

```{r}
plot_grid(p1,p3,p5,p6, ncol = 3, nrow = 2)
```

# 2. Characteristics and Features of an Employee?


## Analyse Attrition vs Years with current manager

### Analyse Feature Years in Current Role verses the Attrition Rate of the employee. We are able to see that employees working with the same manager for 1-9 years have the higest attrition rate of 15.5%

### YearsWithCurrManager : 96.4%

```{r YearsWithCurrManager vs Attrition}
dtCurrentManager <- employeedata%>%
  dplyr::group_by(YearsWithCurrManager,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p8<-ggplot(data = dtCurrentManager,aes(x= YearsWithCurrManager , y = n,fill = Attrition))+ 
   geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Years with Current Manager -Attrition Analysis")+ylab("Count Of Employees")
 
 p8
```
## Analyse Attrition vs Years in Current Role

### Analyse Feature Years in Current Role verses the Attrition Rate of the employee. We are able to see that employees working with the same manager for 1-9 years have the higest attrition rate of 15.4%
```{r}
dtCurrentRole <- employeedata%>%
  dplyr::group_by(YearsInCurrentRole,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p9<- ggplot(data = dtCurrentRole,aes(x= YearsInCurrentRole , y = n,fill = Attrition))+ 
   geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Years In Current Role-Attrition Analysis")+ylab("Count Of Employees")
 
p9
```

## Attrition vs Age of the employee
### Upon analysing emplyees from the age from 30-39 years have higest attrition of 14.2%
### 30-39 years : 38.6%


```{r}
dtAge <- employeedata%>%
  dplyr::group_by(Age,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p10<- ggplot(data = dtAge,aes(x= Age, y = n,fill = Attrition))+ 
   geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Age-Attrition Analysis")+ylab("Count Of Employees")
 
p10
```


## Employee Attrition vs Gender of the employee
```{r}
dtGender <- employeedata%>%
  dplyr::group_by(Gender,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p11<-ggplot(data = dtGender,aes(x=Gender , y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Gender-Attrition Analysis")+ylab("Count Of Employees")

 p11
```


## Employee Attrition vs Martial Status

```{r}
dtMaritial <- employeedata%>%
  dplyr::group_by(MaritalStatus,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p12<-ggplot(data = dtMaritial,aes(x=MaritalStatus, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee MAritial Status-Attrition Analysis")+ylab("Count Of Employees")

 p12

```

## Employee Attrition vs Total Working Years

```{r}
dtWork <- employeedata%>%
  dplyr::group_by(TotalWorkingYears,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p13<-ggplot(data = dtWork ,aes(x= TotalWorkingYears
, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Total Working Years -Attrition Analysis")+ylab("Count Of Employees")

p13
```

## Employee Attrition vs Department
## Most of the employees from the sales department have the highest attrition rate which would be 21.6%

```{r}
  dtDep <- employeedata%>%
    dplyr::group_by(Department,Attrition)%>%
    dplyr::tally()%>%
    dplyr::mutate(percent=n/870)
  
  
  p14<- ggplot(data = dtDep,aes(x= Department, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Department -Attrition Analysis")+ylab("Count Of Employees")

  p14

```

## Education vs the Attrition
## Education level of an employee vs Attrition
```{r}
dtEduc <- employeedata%>%
  dplyr::group_by(Education,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p15<-ggplot(data = dtEduc ,aes(x=Education, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Education Level Option -Attrition Analysis")+ylab("Count Of Employees")

 p15
```

```{r}
dtEducField <- employeedata%>%
  dplyr::group_by(EducationField,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p16<-ggplot(data = dtEducField ,aes(x=EducationField, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Education Level Option -Attrition Analysis")+ylab("Count Of Employees")

 p16
```


## Employee Attrition vs YEarsAtCompany
```{r}
dtYAtCompany <- employeedata%>%
  dplyr::group_by(YearsAtCompany,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p17<- ggplot(data = dtYAtCompany,aes(x= YearsAtCompany, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee Job Level -Attrition Analysis")+ylab("Count Of Employees")

p17

```


## Employee Attrition vs "YearsSinceLastPromotion"

```{r}
dtPromo <- employeedata%>%
  dplyr::group_by(YearsSinceLastPromotion,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p18<- ggplot(data = dtPromo ,aes(x= YearsSinceLastPromotion, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee YEars Since Last Promotion -Attrition Analysis")+ylab("Count Of Employees")

p18

```


## Employee Attrition vs ""RelationshipSatisfaction"
"

```{r}
dtRelationshipsatisfaction <- employeedata%>%
  dplyr::group_by(RelationshipSatisfaction,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/140)


p19<- ggplot(data = dtRelationshipsatisfaction ,aes(x= RelationshipSatisfaction, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee YEars Since Last Promotion -Attrition Analysis")+ylab("Count Of Employees")

p19

```

```{r}
plot_grid(p8,p9,p10,p10,p11,p18,ncol = 3, nrow = 2)
```


## 3. What are relationship between Employee Attrition and the Employee work environment?



## Employee Attrition vs Performance Rating
## Most of the employees from the sales department have the highest attrition rate which would be 21.6%

```{r}
dtPerformance <- employeedata%>%
  dplyr::group_by(PerformanceRating,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p23<-ggplot(data = dtPerformance,aes(x= PerformanceRating
, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Performance Rating -Attrition Analysis")+ylab("Count Of Employees")
p23

```


## Employee Attrition vs Overtime of the Employee

```{r}
dtOver <- employeedata%>%
  dplyr::group_by(OverTime,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p24<-ggplot(data = dtOver ,aes(x=OverTime, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" OVerTime -Attrition Analysis")+ylab("Count Of Employees")
p24
```


## Employee Attrition vs Job Involvement

```{r}
dtJobInv <- employeedata%>%
  dplyr::group_by(JobInvolvement,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p25<-ggplot(data = dtJobInv ,aes(x=JobInvolvement , y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))

p25
```
## Employee Attrition vs Bussiness Travel 
```{r}
dtbus <- employeedata%>%
  dplyr::group_by(BusinessTravel,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/140)


 p26<-ggplot(data = dtbus,aes(x=BusinessTravel , y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Business Travel -Attrition Analysis")+ylab("Count Of Employees")

p26
```



## Employee Attrition vs Distance From Home


```{r}
dtdist <- employeedata%>%
  dplyr::group_by(DistanceFromHome,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p27<-ggplot(data = dtdist,aes(x=DistanceFromHome, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))
 
 p27


```

## Employee Attrition vs JobRole


```{r}
dtJobRole <- employeedata%>%
  dplyr::group_by(JobRole,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/140)


 p28<-ggplot(data = dtJobRole,aes(x=JobRole, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))

 p28

```



## Employee Attrition vs JobLevel


```{r}
dtJoblevel <- employeedata%>%
  dplyr::group_by(JobLevel,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p29<- ggplot(data = dtJoblevel,aes(x=JobLevel, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))
p29

```


## Employee Attrition vs EnvironmentSatisfaction

### Environment Satisfaction
# Environment Satisfaction might be affecting attrition as the lower the environment satisfaction rating, the higher the attrition rate.

```{r}
dtEnv <- employeedata%>%
  dplyr::group_by(EnvironmentSatisfaction,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p30<-ggplot(data = dtEnv ,aes(x=EnvironmentSatisfaction, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))

p30
```

## Employee Attrition vs Training Times Last year

### Training Times Last year vs Employee Attrition


### Employees with Training times less than 2 have more attrition  
```{r}
dttrain <- employeedata%>%
  dplyr::group_by(TrainingTimesLastYear,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


 p31<-ggplot(data = dttrain ,aes(x=TrainingTimesLastYear, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))
p31

```
## Employee Attrition vs "WorkLifeBalance"

### Employees with Training times less than 2 have more attrition  
```{r}
dtWork<- employeedata%>%
  dplyr::group_by(WorkLifeBalance,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/140)


 p32<-ggplot(data = dttrain ,aes(x=TrainingTimesLastYear, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))

p32
```

## Employee Attrition vs worklife balance 

```{r}
dtWorkLife <- employeedata%>%
  dplyr::group_by(WorkLifeBalance,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p33<-ggplot(data =dtWorkLife,aes(x= WorkLifeBalance, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Employee WorkLife BAlance -Attrition Analysis")+ylab("Count Of Employees")

p33
```


## Employee Attrition vs Number of Companies Worked 

```{r}
dtCompanies <- employeedata%>%
  dplyr::group_by(NumCompaniesWorked,Attrition)%>%
  dplyr::tally()%>%
  dplyr::mutate(percent=n/870)


p34<-ggplot(data =dtCompanies,aes(x= NumCompaniesWorked, y = n,fill = Attrition))+ geom_bar(stat="identity")+geom_text(aes(label=paste0(sprintf("%1.1f", percent*100),"%")), position=position_stack(vjust=0.5), colour="white", size = 4)+theme(axis.text.x = element_text(angle = 90,hjust =0 ))+ggtitle(" Number of  Companies Worked -Attrition Analysis")+ylab("Count Of Employees")

p34
```



```{r}
plot_grid(p24,p26,p28,p29,p30,p31,p32,p33,p34,ncol = 3, nrow = 2)
```


## Based on the above analysis we are going to implement KNN model to predict Attrition 


```{r, include=FALSE}
# Change all character variables to factor
employee_updated<-employee
employee_updated$BusinessTravel <- as.integer(as.factor(employee_updated$BusinessTravel))
employee_updated$Department<- as.integer(as.factor(employee_updated$Department))
employee_updated$EducationField <- as.integer(as.factor(employee_updated$EducationField))
employee_updated$Gender <- as.integer(as.factor(employee_updated$Gender))
employee_updated$JobRole <- as.integer(as.factor(employee_updated$JobRole))
employee_updated$MaritalStatus <- as.integer(as.factor(employee_updated$MaritalStatus))
employee_updated$OverTime <- as.integer(as.factor(employee_updated$OverTime))
employee_updated$StockOptionLevel<- as.integer(as.factor(employee_updated$StockOptionLevel))

employee_updated$JobLevel<- as.integer(as.factor(employee_updated$JobLevel))
```

Apply the three main factors with kNN technique to create an Attrition Prediction model

Upon implementing the KNN model to predict attrition we were able to get 82% accuracy and  low pvalue 1.265e-07
## Because of the accuracy we are not using this model to predict Attrition

```{r}
set.seed(8)
splitPerc = .7
KNNdata <- subset(employee_updated, select = c(Attrition,StockOptionLevel, OverTime,JobLevel,Gender,JobRole,Department,MaritalStatus))

trainIndices = sample(1:dim(KNNdata)[1], round(splitPerc * dim(KNNdata)[1]))
train.Attrition = KNNdata[trainIndices,]
test.Attrition = KNNdata[-trainIndices,]
# run knn
classifications = knn(train.Attrition[,c(2:8)], test.Attrition[,c(2:8)], train.Attrition$Attrition, prob = TRUE, k = 21)
# Confusion Matrix
confusionMatrix(table(test.Attrition$Attrition, classifications))
# Find Sensitivity and Specificity for Train Set
classifications_train = knn(train.Attrition[,c(2:8)], train.Attrition[,c(2:8)], train.Attrition$Attrition, prob = TRUE, k = 5)
confusionMatrix(table(train.Attrition$Attrition, classifications_train))
```

Loop for many k and one training / test partition
```{r}
accs = data.frame(accuracy = numeric(30), k = numeric(30))
for(i in 1:30)
{
  classifications = knn(train.Attrition[,c(2:6)], test.Attrition[,c(2:6)], train.Attrition$Attrition, prob = TRUE, k = i)
  table(test.Attrition$Attrition, classifications)
  CM = confusionMatrix(table(test.Attrition$Attrition, classifications))
  accs$accuracy[i] = CM$overall[1]
  accs$k[i] = i
}
plot(accs$k,accs$accuracy, type = "l", xlab = "k")
```
## Leave one out to predict 
```{r, include=FALSE}
# Loop for many k and the average of many training / test partition
iterations = 500
numks = 30
masterAcc = matrix(nrow = iterations, ncol = numks)
   masterSensitivity= matrix(nrow = iterations, ncol = numks)
  masterSpecificity= matrix(nrow = iterations, ncol = numks)
for(j in 1:iterations)
{
accs = data.frame(accuracy = numeric(30), k = numeric(30))
trainIndices = sample(1:dim(KNNdata)[1], round(splitPerc * dim(KNNdata)[1]))
train.Attrition = KNNdata[trainIndices,]
test.Attrition = KNNdata[-trainIndices,]
for(i in 1:numks)
{
  classifications = knn(train.Attrition[,c(2:6)], test.Attrition[,c(2:6)], train.Attrition$Attrition, prob = TRUE, k = i)
  table(classifications,test.Attrition$Attrition)
  CM = confusionMatrix(table(test.Attrition$Attrition, classifications))
  masterAcc[j,i] = CM$overall[1]
  masterSensitivity[j,i]=CM$byClass[1]
   masterSpecificity[j,i]=CM$byClass[2]
}
}
MeanAcc = colMeans(masterAcc)
MeanSensitivity=colMeans(masterSensitivity)
MeanSpecificity=colMeans(masterSpecificity)
plot(seq(1,numks,1),MeanAcc,type = "l")

plot(seq(1,numks,1), MeanSpecificity,type = "l")

plot(seq(1,numks,1),MeanSensitivity,type = "l")
```


```{r}
employeeRF<-employee_updated
employeeRF$Attrition <- factor(employeeRF$Attrition,levels = c('Yes','No'), labels = c('Yes','No'))
set.seed(8)
splitPerc = .7

trainIndices = sample(1:dim(employeeRF)[1], round(splitPerc * dim(employeeRF)[1]))
train.Attrition = employeeRF[trainIndices,]
test.Attrition = employeeRF[-trainIndices,]


## Upon usig the below sample size we were able to get Accuracy 81%,Sensitivity 72% and Specificity 83%
employeeRFmodel<-randomForest(Attrition ~ .,data=train.Attrition, strata=train.Attrition$Attrition,sampsize=c(60,60))


rfpredict <- predict(employeeRFmodel , test.Attrition)

confusionMatrix.random.Forrest <- confusionMatrix(rfpredict, test.Attrition $Attrition)
confusionMatrix.random.Forrest

varImp(employeeRFmodel)


varImpPlot(employeeRFmodel)


##employeeRFmodel1<-randomForest(Attrition ~ .-c(MonthlyRate),data=train.Attrition, strata=train.Attrition$Attrition,sampsize=c(90,90))



#varImpPlot(employeeRFmodel1)
#rfpredict1 <- predict(employeeRFmodel1,test.Attrition)

#confusionMatrix.random.Forrest <- confusionMatrix(rfpredict1,test.Attrition$Attrition)
#confusionMatrix.random.Forrest




```


## EDA for Employee monthly Income Analysis vs Features 


```{r}
employeemonthly<-employee


# 
# ggplot(data=employeemonthly,aes(x=MonthlyIncome,y=Age))+geom_point(color="blue")+ggtitle("Employee Monthly Income vs Age")
# 
# 
# ggplot(data=employeemonthly,aes(x=MonthlyIncome,y=TotalWorkingYears))+geom_point(color="red")+ggtitle("Employee Monthly Income vs Years Of Experience")
# 
# 
# ggplot(data=employeemonthly,aes(x=MonthlyIncome,y=YearsAtCompany))+geom_point(color="green")+ggtitle("Employee Monthly Income vs Years At Company")
# 
# 
# 
# ggplot(data=employeemonthly,aes(x=MonthlyIncome,y=JobLevel))+geom_point(color="plum")+ggtitle("Employee Monthly Income vs Job Level")
# 
# 
# ggplot(employeemonthly,aes(y=employeemonthly,x=Age,color=TotalWorkingYears))+geom_point()
# 
# 
# plot_grid(s1,s2,s3,s4,ncol=2,nrow=2)




```

## Random Forest for Employee Attrition Prediction

```{r}
employeemonthly<-employee_updated



set.seed(8)
splitPerc = .7

trainIndices = sample(1:dim(employeemonthly)[1], round(splitPerc * dim(employeemonthly)[1]))
train.Income = employeemonthly[trainIndices,]
test.Income = employeemonthly[-trainIndices,]



employeeSalaryPred<-lm(MonthlyIncome ~ Age+TotalWorkingYears+YearsAtCompany+JobLevel,data=train.Income)


PredictSal<-predict(employeeSalaryPred,test.Income)

table(RMSE(test.Income$MonthlyIncome,PredictSal))

plot(employeeSalaryPred)

plot(test.Income$MonthlyIncome,PredictSal,type="p",main="Monthly Income Prediction")



plot(test.Income$MonthlyIncome)



```



## Write attrition predicted values for 'CaseStudy2CompSet No Attrition.csv' to csv File
```{r}
# Read in 'No Attrition' file


employeeRF1<-employee_updated
employeeNoAttrition

employeeRF1$Attrition<-as.factor(employeeRF1$Attrition)

rfPredictOriginal=randomForest(Attrition ~ .,data=employeeRF1,strata=employee$Attrition,sampsize=c(60,60))

pred<-predict(rfPredictOriginal,employeeNoAttrition)

Attrition_Pred<-as.data.frame(pred)


ID<-as.data.frame(employeeNoAttrition$ID)

Data_Updated<-cbind(Attrition_Pred,ID)


Attrition_Pred_updated<- merge(x = employeeNoAttrition, y = Data_Updated,
                          by.x = "ID", by.y = "employeeNoAttrition$ID", all = TRUE)


Attrition_Pred_updated$pred

colnames(Attrition_Pred_updated)[36] <- "Attrition"

# Write to file
write.csv(Attrition_Pred_updated, "case_Study_Updated_Attrition_output", row.names = FALSE, quote=FALSE)
```



## Write attrition predicted values for 'CaseStudy2CompSet No Attrition.csv' to csv File

Here we have used the Random Forrest model to predict the attrition of the employeess
```{r}
# Read in 'No Attrition' file


employeeRF1<-employee_updated
employeeNoAttrition

employeeRF1$Attrition<-as.factor(employeeRF1$Attrition)

rfPredictOriginal=randomForest(Attrition ~ .,data=employeeRF1,strata=employee$Attrition,sampsize=c(60,60))

pred<-predict(rfPredictOriginal,employeeNoAttrition)

Attrition_Pred<-as.data.frame(pred)
cbind(Attrition_Pred,ID)

ID<-as.data.frame(employeeNoAttrition$ID)

Data_Updated<-mutate(Attrition_Pred,ID)

Attrition_Pred_updated<- merge(x = employeeNoAttrition, y = Data_Updated,
                          by.x = "ID", by.y = "employeeNoAttrition$ID", all = TRUE)


Attrition_Pred_updated$pred

colnames(Attrition_Pred_updated)[36] <- "Attrition"

# Write to file
write.csv(Attrition_Pred_updated, "Case2PredictionsPearly_PaulAttrition.csv", row.names = FALSE, quote=FALSE)
```


## Write predicted salary values to a csv.

Here we have used Multiple Linear regression to predict the salary of the employee
```{r}
# Read in 'No Attrition' file


employeelinear<-employee_updated
employeeNoSalary

employeelinear$Attrition<-as.factor(employeelinear$Attrition)

employeeSalaryPred1<-lm(MonthlyIncome ~ Age+TotalWorkingYears+YearsAtCompany+JobLevel,data=employeelinear)


PredictSalary<-predict(employeeSalaryPred1,employeeNoSalary)

SalaryPred<-as.data.frame(PredictSalary)

ID_updated<-as.data.frame(employeeNoSalary$ID)

Salary_pred_updated<-cbind(SalaryPred,ID_updated)



Salary_Pred_updated<- merge(x = Salary_pred_updated, y = employeeNoSalary,
                          by.x = "employeeNoSalary$ID", by.y = "ID", all = TRUE)



colnames(Salary_Pred_updated)[2] <- "Monthly_Income"

# Write to file
write.csv(Salary_Pred_updated,"Case2PredictionsPearlyPaulSalary.csv",row.names = FALSE, quote=FALSE)
```

## Conclusion


### Age 
Attrition Rate is higest among people who belong to  the age group of 30 - 39 years
By this we can conclude that employees who have maximum attrition rate are young

### Total Working Years
Employees who have less total working experience between 1-9 years have higher attrition rate
HR needs to overlook why young employees have higher attrition rate.


#### Employees' benefits ####

Salary plays a major influencing factor to attrition
Salary scale varies with the job level


### Stock Option Levels
Employees who  dont receive stocks have the higer attrition rate . Providing stocks from the company employees are forced to stay .So providing Stock level options to junior employees would help.


#### Employees' situations in the company ####

### Overtime
Overtime is the most influential factor of attrition. Around 75% of the employees work overtime. Employees who work overtime tend to quite.
Manager should be adjusting the work time so that employees are not overloaded with work.


### Job Roles
We were able to observe that half of the employees as Lab Technicians were quiting. Possible due to the over time.

### Overtime & Department & JobLevel & JobRole & BusinessTravel

People who work overtime have the highest Attrition rate and belong to the Research and Sales Department. These employees travel rarely and work low level jobs

### Years at the Company 
Employees with higher attrition rate work in the company for around 1-9 years

### Years in Current Role
Most of the employees who leave the company has very less experience in the current role



#### Environment Satisfaction 
Employees with low environment satisfication have higher attrition rate.



### Most Factors that impact Attrition

 1) Employees’ background 
 Working experience
## 2) Employees’ working Environment
a) Working overtime
 
c) Job levels
3) Employees’ benefits
  a) Salary
 b) Stock options


## RECOMMENDATIONS FROM DDS ANALYTICS :

1) Career counseling and coaching should be provided for new employees
2) Junior Employees should be getting Stock Options as part of their salary package.
3) Employees should not be overloaded by extra work that they are forced to do Overtime
4) Company Promotion criteria should be reviewed regularly at least  on a yearly basis.
5) Working environment should be improved . Employees should be able to work were they feel comfortable whether it is the office or at home.
6)We would recommend HR to analyze the Employees in the Research/Sales Department to check if there are any environment issues or work issues that is causing employees to leave the company.



```{r}

```